---
title: "Time Series and Forecasting - Project"
author: 
  - "Apoorva Prakash"
  - "Halleluya Mengesha"
  - "Hira Stanley"
date: "2025-02-01"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

# Data Loading and Cleaning

Lorem Ipsum

**Lorem Ipsum **

### Install Required Libraries

```{r, message=FALSE}

# Install/load needed packages
# install.packages("lubridate")
# install.packages("forecast")     
# install.packages("fpp3")         
library(fpp3)
library(tsibble)
library(forecast)
library(lubridate)
library(fable)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tseries)

```

### Load Dataset

```{r, message=FALSE, warning=FALSE}

df <- read.csv("C:/Users/apoor/Downloads/aqi_exo.csv")

```

### Initial Inpection

```{r, message=FALSE, warning=FALSE}

glimpse(df)

```

```{r, message=FALSE, warning=FALSE}

head(df)

```

```{r, message=FALSE, warning=FALSE}

# Quick summary stats
summary(df)

```

### Clean the data

```{r, message=FALSE, warning=FALSE}

# Convert Date to Date format
df <- df %>%
  mutate(Date = ymd(Date)) %>%
  arrange(Date)

# Convert pollutant columns to numeric 
# Handle missing values
df <- df %>%
  mutate(across(c("CO", "Ozone", "PM10", "PM25", "NO2"), 
                ~ as.numeric(na_if(na_if(., "."), ""))))

# Rename columns
colnames(df) <- c("date", "aqi", "main_pollutant", "site_name", "site_id", "source", 
                  "co", "ozone", "pm10", "pm25", "nO2", "avg_temp", "avg_humidity",
                  "avg_wind_speed")

```

```{r, message=FALSE, warning=FALSE}

# Check missing values
(colSums(is.na(df)))

print(paste("Total missing values is: ", sum(is.na(df))))

```

```{r, message=FALSE, warning=FALSE}

# Convert to a tsibble
ts_data <- df %>%
  select(date, aqi, co, ozone, pm10, pm25, nO2, avg_temp, avg_humidity, avg_wind_speed) %>%
  # filter(year(date) >= 2010) %>% 
  as_tsibble(index = date)

```


### Plot the data

#### Overall time series
```{r, message=FALSE, warning=FALSE}

# Plot AQI  over time
ts_data %>%
  autoplot(aqi) +
  labs(title = "NYC Air Quality Index (AQI) Over Time",
       x = "Date", y = "AQI Value") +
  theme_minimal()

```

```{r, message=FALSE, warning=FALSE}

# Plot all pollutants over time (Facet Grid)
ts_data %>%
  pivot_longer(cols = c(co, ozone, pm10, pm25, nO2), names_to = "Pollutant", values_to = "Value") %>%
  ggplot(aes(x = date, y = Value, color = Pollutant)) +
  geom_line() +
  facet_wrap(~Pollutant, scales = "free_y") +
  labs(title = "Pollutant Trends Over Time", x = "Date", y = "Pollutant Level") +
  theme_minimal()

```

```{r, message=FALSE, warning=FALSE}
# Plot AQI  over time - 2023
ts_data %>%
  filter(year(date) == 2023) %>% 
  autoplot(aqi) +
  labs(title = "NYC Air Quality Index (AQI) Over Time - 2023",
       x = "Date", y = "AQI Value") +
  theme_minimal()
```

#### Decomposition

```{r, message=FALSE, warning=FALSE}

# Check AQI - seasonal trends
ts_data %>%
  # filter(year(date) >= 2022) %>%
  gg_season(aqi) +
  labs(title = "Seasonal Pattern of AQI", x = "Month", y = "AQI Value")

```

```{r, message=FALSE, warning=FALSE}

# Check AQI - seasonal trends - 2022 to 2024
ts_data %>%
  filter(year(date) >= 2020) %>%
  gg_season(aqi) +
  labs(title = "Seasonal Pattern of AQI: 2020 to 2024", x = "Month", y = "AQI Value")

```

```{r, message=FALSE, warning=FALSE}

# Classical decomposition for AQI
aqi_decomp <- ts_data %>%
  # filter(year(date) >= 2023) %>%
  # filter(year(Date) == 2023) %>%
  model(classical_decomposition(aqi, type = "additive")) %>%
  components()

# Plot decomposed components
aqi_decomp %>%
  autoplot() +
  labs(title = "Classical Decomposition of AQI Time Series- 2023 to 2024")

```


```{r, message=FALSE, warning=FALSE}

# STL(Seasonal-Trend Decomposition using Loess) - for AQI
ts_data %>%
  # filter(year(date) >= 2019) %>%
  model(STL(aqi ~ season(window = "periodic"))) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition of AQI Time Series")

```

#### Augmented Dickey-Fuller (ADF) test

```{r, message=FALSE, warning=FALSE}

# Perform the ADF test
adf_result <- adf.test(ts_data$aqi, alternative = "stationary")
print(adf_result)

```

```{r, message=FALSE, warning=FALSE}

# ACF plot to check autocorrelation
ts_data %>%
  ACF(`aqi`) %>%
  autoplot() +
  labs(title = "ACF Plot of NYC AQI",
       x = "Lag", y = "Autocorrelation")

# PACF plot to check partial autocorrelation
ts_data %>%
  PACF(`aqi`) %>%
  autoplot() +
  labs(title = "PACF Plot of NYC AQI",
       x = "Lag", y = "Partial Autocorrelation")


```

### Check Stationarity

**<u>Observations</u>**:
\

**1. Trend**: 

* "NYC Air Quality Index (AQI) Over Time" plot shows a slightly decreasing trend over time.
* This suggests the mean and the variance are not constant.
\

→ **Data requires differencing**. We should use trend-based model (**ARIMA** with differencing)
\

→ **Non-Seasonal Differencing term: d = 1**
\

**2. Seasonality**: 

* "NYC Air Quality Index (AQI) Over Time" plot shows some seasonality
* "Seasonal Trend Decomposition for AQI" plot shows a strong seasonal component with AQI values repeating periodically.
\

→ **We should use seasonal differencing approach** (**SARIMA**)
\

→ **Seasonal Differencing : D = 1, S = 3**
\

**3. ACF**: 

* Repeated spikes every 3 days are present. This means 3-day cycle/seasonality is present possibly due to external factors)
* Gradual decline can be seen. This confirms trend effect.
\

→ **MA (q, Q)**: **q = 1, Q = 1** (q from ACF cutoff and, Q represents one past seasonal MA term included at lag 3)

**4. PACF**: 

* High PCAF at Lag 1 - AQI today is strongly correlated with AQI yesterday (Autoregressive Process AR(1))
* Spikes every 3 days - Seasonality
\

→ **AR (p, P)**: **p = 1, P = 1** (from ACF cutoff and, P represents one past seasonal AR term is included at lag 3)
\

**<u>Conclusion</u>: Data is non-stationary despite the ADF test since it does not take into account strong trend and seasonal patterns.**

\newpage

### Fit Models and Forecast

#### Fit Baseline Models (Naive & S-Naive)

* **Note**: Simple models as a benchmark: to see if complex models outperform simple predictions.

```{r}
#Splitting into train and test (year 2024) data
train_data <- ts_data %>% filter(year(date) < 2024)
test_data <- ts_data %>% filter(year(date) == 2024)
```


```{r, message=FALSE, warning=FALSE}


# Naïve model 
naive_model <- train_data %>%
  model(NAIVE(aqi))

# Report

report(naive_model)

```
```{r, message=FALSE, warning=FALSE}

# Seasonal Naïve Model 
snaive_model <- train_data %>%
  model(SNAIVE(aqi ~ lag("3 days")))

# Report
report(snaive_model)

```

```{r, message=FALSE, warning=FALSE}

# Filter data for plot purposes
ts_data_filtered <- train_data %>%
  filter(year(date) > 2020)

# Forecast for the year 2024
naive_forecast <- forecast(naive_model, new_data = test_data)
snaive_forecast <- forecast(snaive_model, new_data = test_data)

# Plot forecasts
autoplot(ts_data_filtered, aqi) +
  autolayer(naive_forecast, series= "Naive Forecast", PI = FALSE) +
  labs(title = "Naïve Forecast (Baseline)")

autoplot(ts_data_filtered, aqi) +
  autolayer(snaive_forecast, series= "Snaive Forecast", PI = FALSE) +
  labs(title = "Seasonal Naïve Forecast (3-Day Cycle)")

```

\

**<u>Observations</u>**:
\

**1. Naive Model (remains the same as last observed value):**

* **No trend** 
* **No seasonality**

→ Very weak baseline model (expected). Only useful for comparison—real AQI data has seasonality & trends.
\

**2. Snaive (repeats last season's value):**

* No trend
* Short-term seasonality

→ Improves naïve model but still too simple.

#### Fit ARIMA/SARIMA 

```{r, message=FALSE, warning=FALSE}

# Fit Auto ARIMA model
auto_arima_model <- train_data %>%
  model(ARIMA(aqi))

# Report
report(auto_arima_model)

```

```{r, message=FALSE, warning=FALSE}

# Forecast for the year 2024
auto_arima_forecast <- forecast(auto_arima_model, new_data=test_data)

# Plot forecast
autoplot (ts_data_filtered, aqi) + 
  autolayer(auto_arima_forecast, series= "Arima Forecast", PI = FALSE) +
  labs(title = "AQI Forecast using Auto ARIMA")

```
\

**<u>Observations</u>**:
\

**ARIMA Model (selects the best ARIMA/SARIMA model using AIC/BIC):**

\

```{r, message=FALSE, warning=FALSE}

gg_tsresiduals(auto_arima_model)

```
\

**<u>Observations</u>**:
\

* Almost all spikes are outside the blue bounds in ACF suggesting potential **missing seasonal effects** or  **exogenous factors **. 
* A more complex model  might improve fit. We will try to manually determine/adjust SARIMA parameters (p, d, q) and (P, D, Q, S).

\
```{r, message=FALSE, warning=FALSE}

sarima_model_manual <- train_data %>%
  model(ARIMA(aqi ~ pdq(1,1,1) + PDQ(1,1,1,3)))  # (S=3 for 3-day seasonality)

report(sarima_model_manual)
```
```{r}

# Forecast using SARIMA
sarima_forecast_manual <- forecast(sarima_model_manual, new_data = test_data)

# Plot the new forecast
autoplot(ts_data_filtered, aqi) + 
  autolayer(sarima_forecast_manual, series= "SARIMA Forecast", PI = TRUE) +
  labs(title = "AQI Forecast using Manual SARIMA (S=3)")

```

```{r}

gg_tsresiduals(sarima_model_manual)

```
\

**<u>Observations</u>**:
\

* The spikes outside the blue bounds in ACF suggest potential **missing seasonal effects** or  **exogenous factors ** (e.g., weather, traffic data).
* We will try  **SARIMAX ** ( **SARIMA  +  exogenous variables **) to account for other AQI influences.

\

```{r}
# Compare model performance

# AIC and BIC separately
aic_bic_tbl <- bind_rows(
  glance(naive_model) %>% mutate(Model = "Naïve"),
  glance(snaive_model) %>% mutate(Model = "S-Naïve"),
  glance(auto_arima_model) %>% mutate(Model = "Auto ARIMA"),
  glance(sarima_model_manual) %>% mutate(Model = "Manual SARIMA")
) %>%
  select(Model, AIC, BIC)

# RMSE and other accuracy metrics
accuracy_tbl <- bind_rows(
  accuracy(naive_model) %>% mutate(Model = "Naïve"),
  accuracy(snaive_model) %>% mutate(Model = "S-Naïve"),
  accuracy(auto_arima_model) %>% mutate(Model = "Auto ARIMA"),
  accuracy(sarima_model_manual) %>% mutate(Model = "Manual SARIMA")
) %>%
  select(Model, ME, RMSE, MAE, MASE)

# Print accuracy and model comparison
print(accuracy_tbl)
print(aic_bic_tbl)
```
\

**<u>Observations</u>**:
\

**Manual SARIMA is the best model:**:

* Lowest RMSE (15.88) → Best predictive accuracy
* Lowest MAE (11.07) → Least forecasting error on average
* Lower AIC → Better model fit (better accuracy)
* Lower BIC → Reduced complexity (penalty on extra parameters)

indicate a better model penalize model complexity, so this suggests SARIMA is both accurate and efficient.

The spikes outside the blue bounds in ACF suggest potential **missing seasonal effects** or  **exogenous factors ** (e.g., weather, traffic data).
* We will try  **SARIMAX ** ( **SARIMA  +  exogenous variables **) to account for other AQI influences.

\


#### Fit SARIMAX

* **Note**: Including wether data (temperature, wind and humidity)

```{r}

# Fit SARIMAX model with exogenous variables
sarimax_model <- train_data %>%
  model(
    SARIMAX = ARIMA(aqi ~ avg_temp + avg_humidity + avg_wind_speed +
                        pdq(1,1,1) + PDQ(1,1,1,3))  # Weekly seasonality
  )

# Print model summary
report(sarimax_model)

```
```{r}
# Forecast using SARIMAX
sarimax_forecast <- forecast(sarimax_model, new_data = test_data)

# Plot the new forecast
autoplot(ts_data_filtered, aqi) + 
autolayer(sarimax_forecast, series= "SARIMAX Forecast", PI = TRUE) +
labs(title = "AQI Forecast using SARIMAX - with Temperature, Humidity and Wind Speed")
```

```{r}
gg_tsresiduals(sarimax_model)
```

```{r}
# Compare model performance

# AIC and BIC separately
aic_bic_tbl <- bind_rows(
  glance(naive_model) %>% mutate(Model = "Naïve"),
  glance(snaive_model) %>% mutate(Model = "S-Naïve"),
  glance(auto_arima_model) %>% mutate(Model = "Auto ARIMA"),
  glance(sarima_model_manual) %>% mutate(Model = "Manual SARIMA"),
  glance(sarimax_model) %>% mutate(Model = "Manual SARIMAX")
) %>%
  select(Model, AIC, BIC)

# RMSE and other accuracy metrics
accuracy_tbl <- bind_rows(
  accuracy(naive_model) %>% mutate(Model = "Naïve"),
  accuracy(snaive_model) %>% mutate(Model = "S-Naïve"),
  accuracy(auto_arima_model) %>% mutate(Model = "Auto ARIMA"),
  accuracy(sarima_model_manual) %>% mutate(Model = "Manual SARIMA"),
  accuracy(sarimax_model) %>% mutate(Model = "SARIMAX")
) %>%
  select(Model, ME, RMSE, MAE, MASE)

# Print accuracy and model comparison
print(accuracy_tbl)
print(aic_bic_tbl)
```
\

**<u>Observations</u>**:
\

**SARIMAX seems to be the best model:**:

* Lowest RMSE (14.63) → Best predictive accuracy
* Lowest MAE (10.32) → Least forecasting error on average
* Lowest MASE (0.56) → Outperforms naïve benchmarkmodels (< 1)
* Minimal Mean Error (0.046) → Lower bias in its predictions
* Lowest AIC → Better model fit (better accuracy)
* Lowest BIC → Reduced complexity (penalty on extra parameters)

\

#### Supplementary Question

* **How do weather conditions affect AQI?**



```{r}
#install.packages("corrr")
library(corrr)

# Compute correlations using corrr package
weather_cor <- ts_data %>%
  select(aqi, avg_temp, avg_humidity, avg_wind_speed) %>%
  correlate()

print(weather_cor)

# View correlations between aqi and other variables
aqi_correlations <- weather_cor %>%
  focus(aqi)

print(aqi_correlations)

ts_data$aqi <- as.numeric(ts_data$aqi) #since it is int and others num

# Regression model: AQI vs. Weather conditions
weather_model <- lm(aqi ~ avg_temp + avg_humidity + avg_wind_speed, data = ts_data)
summary(weather_model)

# Plot residuals to check assumptions
plot(weather_model)
```

\

**<u>Observations</u>**:
\


* Wind Speed has the strongest negative correlation (-0.28) - Higher wind speeds likely help disperse pollutants, reducing AQI levels.
* Temperature has a weak positive correlation (0.086) - Warmer temperatures might contribute to higher AQI due to increased chemical reactions forming secondary pollutants (e.g., ozone).
* Humidity shows minimal correlation (-0.008) - While humidity can impact certain pollutants (e.g., particulate matter), its overall effect on AQI in this dataset is weak.

\


#### ~ End of Document ~




